Distinguishing multiple-merger from Kingman coalescence using two-site frequency spectra

### 2018-10-05 manuscript/manuscript.tex
Ivana and Michael suggestions (part 1).

An important application of the Kingman coalescent is inferring historical population sizes from genetic data~\autocite{SchraiberAkey2015}.
In its simplest form, the model has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982}.
Under a variety of conditions, the coalescent rate is inversely proportional to the population size~\autocite{Kingman1982a}.
Accordingly, a growing or shrinking population may be modeled by a time-varying rate~\autocite{GriffithsTavare1994, GriffithsTavare1998}.
Patterns of genetic diversity depend on the ratio of the coalescent rate to other evolutionary rate parameters.
For example, the \emph{site frequency spectrum} (SFS)---the number of mutations segregating at different frequencies in a sample---is determined by the ratio of the mutation rate to the (time-varying) coalescent rate.
Kingman coalescent--based inference methods solve the inverse problem of determining the population size history that best explains particular features of the data, such as the SFS (e.g., \cite{BhaskarEtAl2015}) or variations in heterozygocity along a chromosome (e.g., \cite{LiDurbin2011}).

@@

While the example above is extreme, it is well-established that violations of the neutrality assumption can distort or mask the signatures of population size changes.
For example, \cite{SchriderEtAl2016} recently demonstrated that several popular inference methods give misleading results in the presence of selective sweeps.
In a similar vein, \cite{CvijovicEtAl2018}, showed that purifying selection at linked sites that is sufficient to reduce genetic diversity is also sufficient to distort the SFS, leading to a false signal of population growth.
Moreover, genomic evidence from multiple species suggests that such violations of neutrality may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.

An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} \autocite{Pitman1999, Sagitov1999, DonnellyKurtz1999,} (reviewed in \cite{Eldon2016}), which arise in a variety of contexts both with and without selection.
Whereas in the Kingman coalescent lineages may coalesce only pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
The even more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one distinct multiple merger event at the same time.
These models are relevant for species with
``sweepstakes'' reproductive events \autocite{EldonWakeley2006, SargsyanWakeley2008},
fat-tailed offspring number distributions \autocite{Schweinsberg2003},
recurring selective sweeps at linked sites \autocite{DurrettSchweinsberg2005, CoopRalph2012},
rapid adaptation \autocite{NeherHallatschek2013, DesaiEtAl2013},
and purifying selection at sufficiently dense sites \autocite{SegerEtAl2010, NicolaisenDesai2012}.

@@

\cite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, varies from 2 to 1.
Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003, SteinruckenEtAl2013}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in rapidly adapting populations \autocite{NeherHallatschek2013}.
Like Fu, Birkner et al.\ were primarily concerned with computing the sample variance of SFS-based summary statistics such as Tajima's D \autocite{Tajima1989}.
As a result, they were mostly interested in the diagonal terms of the SFS covariance matrix, which dominate that calculation.
However, Figures 5 and 6 of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \not\in \{i, n-i\}$.
Thus, unlike the standard Kingman coalescent, the beta coalescent can generate positive associations between mutations with different minor allele counts.

@@

We claim that the second benefit of computing $\fpmi$ from the 2-SFS is that it normalizes for distortions in the coalescent time distribution.
Population-size variation distorts the SFS by changing the distribution of waiting times.
However, the time-inhomogeneous Kingman coalescent generates the same distribution of tree topologies as the time-homogeneous version.
On the other hand, multiple mergers alter both the coalescent time distribution and the distribution of topologies.
These two effects are convolved in the SFS, making it difficult to distinguish between population growth and multiple mergers.
Our hope is that the $\fpmi$ reflects additional information about the distribution of genealogies, once distortions in coalescent times are accounted for.
That is, we expect $\fpmi$ to be sensitive to multiple mergers, but invariant to changes in the coalescent rate.

@@

For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with length $d$ and per basepair recombination rate $r$ resulting in a per locus total recombination rate $d r$.
(See SI for parameter combinations.)
We chose values of $d r$ so that the realized values of $d r \E{T_2}$ varied over several orders of magnitude.
For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at each end of the loci.
We then calculated the expectations $\{\E{\tau_i}\}$ and $\{\E{\tau_{ij}}\}$ by averaging over independent loci.
These expectations allow us to calculate all of the 2-SFS statistics defined above.

@@

We simulated a model of recurring selective sweeps using the software \texttt{SLiM} \autocite{Messer2013}.
In all of these simulations, we simulated a population of 500 diploids for $10^4$ generations.
Each haploid genome consisted of a single genomic element $L=10^8$ basepairs long with recombination rate per basepair $r = 10^{-8}$ and overall mutation rate $\mu = 10^{-7}$.
We simulated two types of mutations: neutral mutations and beneficial mutations with additive effects and selection coefficient $s=0.1$.
With these parameters, $2Ns = 100$, so beneficial mutations are strongly selected and will sweep in $T_{\text{sweep}}\sim s^{-1} \log Ns \approx 50$ generations.
Such sweeps will effect a region of the chromosome $d_{\text{sweep}} \sim (r T_{\text{sweep}})^{-1} \approx 2 \times 10^6$ basepairs long.
Thus $d_{\text{sweep}} \ll L$, which will minimize edge effects of simulating a finite chromosome.

@@

The previous sections have shown that $\fpmi$ between pairs of non-recombining sites can discriminate between the Kingman coalescent with population growth and the beta coalescent.
However, most demographic inference is performed on regions of the genome with non-zero recombination rates.
The advantage of the $\fpmi$ approach is that it can be computed on a genomic scale, as with the SFS.
Therefore, it is important to assess the robustness of our approach to recombination between sites.

@@

\Fig{fig:wfpmi_recombination} shows the weighted $\fpmi$ for three genetic distances in a constant-$N$ Kingman coalescent, a Kingman coalescent with exponential growth, and a beta coalescent.
As in the non-recombining case, the constant-$N$ and exponential-growth Kingman models have similar $\fpmi$.
In both models, $\fpmi > 0$ for $i\approx j$ for $d r \E{T_2} \sim 1$.
This is presumably because trees at nearby sites contain clades with similar numbers of leaves.
These positive correlations do not extend to $i \gg j$, which is the signal of multiple mergers in our binned $\hilopmi$.
On the other hand, the positive $\fpmi$ in the beta coalescent persists for $d r \E{T_2} > 1$.
Thus, the signal of multiple mergers in $\fpmi$ is robust to recombination.

@@

\Fig{fig:hilo_vs_d} shows $\hilopmi(d,i_c)$ in both models of growth and the beta coalescent, for a range of $d r \E{T_2}$ and three different choices of $i_c$.
Each curve represents a particular parameter combination and coalescent model and is colored by the distortion in the average SFS, which is independent of the recombination rate.
At low recombination rates, $\hilopmi$ may be greater in growing populations than in constant-$N$ populations (dotted lines), but is always less than zero, consistent with the results for non-recombining sites.
When $d r \E{T_2}\geq 1$, $\hilopmi$ may be slightly positive for large $i_c$, but is smaller in growing populations than in the constant-$N$ Kingman model.
In all Kingman models, $\hilopmi$ decays to zero for $d r \E{T_2}\gg 1$.

@@

\Fig{fig:hilo_vs_tajD} shows the relative invariance of $\hilopmi$ to a time-varying coalescent rate.
While population growth in a Kingman coalescent model can strongly distort the SFS, as measured by Tajima's D, it has very little effect on the $\hilopmi$ for sites $d r \E{T_2} \sim 1$ apart.
This holds for exponential as well as piecewise constant growth.
In contrast, the beta coalescent induces large positive $\hilopmi$ for the same values of Tajima's D.
The figure also showes that this result is robust to the choice of cutoff minor allele count for binning.

\begin{figure}
\centering
\includegraphics[scale=1]{figures/hilopmi_vs_tajimasD_recombination.pdf}
\caption{Pointwise mutual information between high-- and low--minor allele count mutations ($\hilopmi$) versus normalized Tajima's D for simulations with recombination. All values are for $d r \E{T_2} \sim 1$ and $n=100$. Orange points show Kingman coalescent simulations with population growth. Blue points show beta coalescent simulations. \label{fig:hilo_vs_tajD}}
\end{figure}

@@

The bottom row of \fig{fig:hilo_vs_d} shows the results of these simulations.
For intermediate genetic distances, $d r \E{T_2} \sim 1$, the effects of linked sweeps are qualitatively similar to the effects of the beta coalescent.
That is, when sweeps are sufficiently frequent to distort the SFS, as measured by Tajima's D, they also increase $\hilopmi$.
The primary difference between the forward-time sweeps and beta coalescent simulations, is that the distortions caused by sweeps decay to zero for $d r \E{T_2} \gg 1$.
This is reasonable because the effect of a particular sweep on genealogies should be localized around the beneficial mutation.

@@

>>>>

An important application of the Kingman coalescent is inferring historical population sizes from genetic data~\autocite{SchraiberAkey2015}.
In its simplest form, the model has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982}.
Under a variety of conditions, the coalescent rate is inversely proportional to the population size~\autocite{Kingman1982a}.
Accordingly, a growing or shrinking population may be modeled by a time-varying rate~\autocite{GriffithsTavare1994, GriffithsTavare1998}.
Patterns of genetic diversity depend on the ratio of the coalescent rate to other evolutionary rate parameters.
For example, the \emph{site frequency spectrum} (SFS)---the number of mutations segregating at different frequencies in a sample---is determined by the ratio of the mutation rate to the (time-varying) coalescent rate.
Kingman coalescent--based inference methods solve the inverse problem of determining the population size history that best explains particular features of the data, such as the SFS (e.g., \cite{BhaskarEtAl2015}) or variations in heterozygosity along a chromosome (e.g., \cite{LiDurbin2011}).

@@

While the example above is extreme, it is well-established that violations of the neutrality assumption can distort or mask the signatures of population size changes.
For example, \cite{SchriderEtAl2016} recently demonstrated that several popular inference methods give misleading results in the presence of selective sweeps.
In a similar vein, \cite{CvijovicEtAl2018}, showed that reduction of genetic diversity by purifying selection is accompanied by distortions in the SFS, leading to a false signal of population growth.
Moreover, genomic evidence from multiple species suggests that such violations of neutrality may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.

An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} \autocite{Pitman1999, Sagitov1999, DonnellyKurtz1999,} (reviewed in \cite{Eldon2016}), which arise in a variety of contexts both with and without selection.
Whereas in the Kingman coalescent lineages may coalesce only pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
The even more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one distinct multiple merger event at the same time.
These models are relevant for species with
``sweepstakes'' reproductive events \autocite{EldonWakeley2006, SargsyanWakeley2008},
fat-tailed offspring number distributions \autocite{Schweinsberg2003},
recurring selective sweeps at linked sites \autocite{DurrettSchweinsberg2005, CoopRalph2012},
rapid adaptation \autocite{NeherHallatschek2013, DesaiEtAl2013},
and purifying selection at sufficiently many sites \autocite{SegerEtAl2010, NicolaisenDesai2012, GoodEtAl2014}.

@@

\cite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, varies from 2 to 1.
Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003, SteinruckenEtAl2013}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in populations that are rapidly adapting or experiencing extensive purifying selection \autocite{NeherHallatschek2013}.
The calculations of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \not\in \{i, n-i\}$ (Figures 5 and 6).
Thus, unlike the standard Kingman coalescent, the beta coalescent can generate positive associations between mutations with different minor allele counts.

@@

We claim that the second benefit of computing $\fpmi$ from the 2-SFS is that it normalizes for distortions in the coalescent time distribution.
Population-size variation distorts the SFS by changing the distribution of waiting times.
However, the time-inhomogeneous Kingman coalescent generates the same distribution of tree topologies as the time-homogeneous version.
On the other hand, multiple mergers alter both the coalescent time distribution and the distribution of topologies.
These two effects are convolved in the SFS, making it difficult to distinguish between population growth and multiple mergers.
Our hope is that the $\fpmi$ reflects additional information about the distribution of genealogies, once distortions in coalescent times are accounted for.
That is, we expect $\fpmi$ to be sensitive to multiple mergers, but relatively unaffected by changes in the coalescent rate.

@@

For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with length $d$ and per basepair recombination rate $r$ resulting in a per locus total recombination rate $d\,r$.
(See SI for parameter combinations.)
We chose values of $d\,r$ so that the realized values of $d\,r \E{T_2}$ varied over several orders of magnitude.
For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at each end of the loci.
We then calculated the expectations $\{\E{\tau_i}\}$ and $\{\E{\tau_{ij}}\}$ by averaging over independent loci.
These expectations allow us to calculate all of the 2-SFS statistics defined above.

@@

We simulated a model of recurring selective sweeps using the software \texttt{SLiM} \autocite{Messer2013}.
In all of these simulations, we simulated a population of 500 diploids for $10^4$ generations.
Each haploid genome consisted of a single genomic element $L=10^8$ basepairs long with recombination rate per basepair $r = 10^{-8}$ and per basepair mutation rate $\mu = 10^{-7}$.
We simulated two types of mutations: neutral mutations and beneficial mutations with additive effects and selection coefficient $s=0.1$.
With these parameters, $2Ns = 100$, so beneficial mutations are strongly selected and will sweep in $T_{\text{sweep}}\sim s^{-1} \log Ns \approx 50$ generations.
Such sweeps will effect a region of the chromosome $d_{\text{sweep}} \sim (r T_{\text{sweep}})^{-1} \approx 2 \times 10^6$ basepairs long.
Thus $d_{\text{sweep}} \ll L$, which will minimize edge effects of simulating a finite chromosome.

@@

We have shown in the previous sections that $\fpmi$ between pairs of non-recombining sites can discriminate between the Kingman coalescent with population growth and the beta coalescent.
However, most demographic inference is performed on regions of the genome with non-zero recombination rates.
The advantage of the $\fpmi$ approach is that it can be computed on a genomic scale, as with the SFS.
Therefore, it is important to assess the robustness of our approach to recombination between sites.

@@

\Fig{fig:wfpmi_recombination} shows the weighted $\fpmi$ for three genetic distances in a constant-$N$ Kingman coalescent, a Kingman coalescent with exponential growth, and a beta coalescent.
As in the non-recombining case, the constant-$N$ and exponential-growth Kingman models have similar $\fpmi$.
In both models, $\fpmi > 0$ for $i\approx j$ for $d\,r \E{T_2} \sim 1$.
This is presumably because trees at nearby sites contain clades with similar numbers of leaves.
These positive correlations do not extend to $i \gg j$, which is the signal of multiple mergers in our binned $\hilopmi$.
On the other hand, the positive $\fpmi$ in the beta coalescent persists for $d\,r \E{T_2} > 1$.
Thus, the signal of multiple mergers in $\fpmi$ is robust to recombination.

@@

\Fig{fig:hilo_vs_d} shows $\hilopmi(d,i_c)$ in both models of growth and the beta coalescent, for a range of $d\,r \E{T_2}$ and three different choices of $i_c$.
Each curve represents a particular parameter combination and coalescent model and is colored by the distortion in the average SFS, which is independent of the recombination rate.
At low recombination rates, $\hilopmi$ may be greater in growing populations than in constant-$N$ populations (dotted lines), but is always less than zero, consistent with the results for non-recombining sites.
When $d\,r \E{T_2}\geq 1$, $\hilopmi$ may be slightly positive for large $i_c$, but is smaller in growing populations than in the constant-$N$ Kingman model.
In all Kingman models, $\hilopmi$ decays to zero for $d\,r \E{T_2}\gg 1$.

@@

\Fig{fig:hilo_vs_tajD} shows the relative invariance of $\hilopmi$ to a time-varying coalescent rate.
While population growth in a Kingman coalescent model can strongly distort the SFS, as measured by Tajima's D, it has very little effect on the $\hilopmi$ for sites $d\,r \E{T_2} \sim 1$ apart.
This holds for exponential as well as piecewise constant growth.
In contrast, the beta coalescent induces large positive $\hilopmi$ for the same values of Tajima's D.
The figure also showes that this result is robust to the choice of cutoff minor allele count for binning.

\begin{figure}
\centering
\includegraphics[scale=1]{figures/hilopmi_vs_tajimasD_recombination.pdf}
\caption{Pointwise mutual information between high-- and low--minor allele count mutations ($\hilopmi$) versus normalized Tajima's D for simulations with recombination. All values are for $d\,r \E{T_2} \sim 1$ and $n=100$. Orange points show Kingman coalescent simulations with population growth. Blue points show beta coalescent simulations. \label{fig:hilo_vs_tajD}}
\end{figure}

@@

The bottom row of \fig{fig:hilo_vs_d} shows the results of these simulations.
For intermediate genetic distances, $d\,r \E{T_2} \sim 1$, the effects of linked sweeps are qualitatively similar to the effects of the beta coalescent.
That is, when sweeps are sufficiently frequent to distort the SFS, as measured by Tajima's D, they also increase $\hilopmi$.
The primary difference between the forward-time sweeps and beta coalescent simulations, is that the distortions caused by sweeps decay to zero for $d\,r \E{T_2} \gg 1$.
This is reasonable because the effect of a particular sweep on genealogies should be localized around the beneficial mutation.

@@

\subsection*{Acknowledgements}
TODO
